---
title: "Docker R Notebook task"
output: html_docker_task
---

Open the list of packages installed:
```{r}
library()
```

Of those, 10 were loaded:
```{r}
search()
```

Understanding the use of vignette() to view package vignettes
```{r}
?vignette
```

Help with installing a package is accessible here:
```{r}
?install.packages
```

Installing seqinr package:
```{r}
install.packages("seqinr",repos="https://cran.rstudio.com/")  
```
For seqinr information:
```{r}
library(help="seqinr") 
```

Loading seqinr in order to list its functions:
```{r}
library(seqinr) 
```

Listing its functions (207 in total):
```{r}
ls("package:seqinr") 
```
Describing the method a in seqinr (converts 3-letter aa code to 1-letter code)
```{r}
?seqinr::a
```
 
Testing for Tyrosine
```{r}
seqinr::a("Tyr")
```
Checking what the method seqinr::words() does:
```{r}
?seqinr::words
```
It returns all words with the size of the first argument made from the alphabet defined by the vector on the second argument

The following creates all the 64 words of size 3 that are possible from the alphabet defined by the vector on the second argument:
```{r}
seqinr::words(3, c("A", "G", "C", "U"))
```

This lists the available datasets:
```{r}
?data
```

This shows, for example, that seqinr contains data with light absobtion of a water column at the dataset called waterabs (last entry)
```{r}
data(package="seqinr")
```

Loading aaindex dataset
```{r}
data(aaindex, package="seqinr")  
```

This dataset contains 544 properties (physical, chemical and biological) for the 20 natural amino-acids
```{r}
?aaindex 
```

For example, a type of hydrophobicity index:
```{r}
aaindex$FASG890101
```
Another example: the average volume of each amino-acid:
```{r}
aaindex$PONJ960101 
```

This will plot hydrophobicity VS volume for each residue
```{r}
plot(aaindex$FASG890101$I,
     aaindex$PONJ960101$I,
     xlab="hydrophobicity", ylab="volume", type="n")
text(aaindex$FASG890101$I,
     aaindex$PONJ960101$I,
     labels=a(names(aaindex$FASG890101$I)))
```
A beautiful plot with the position of some types of amino-acids (e.g. acidic residues) in the MBP1 sequence from yeast, and the amount of each residue.
```{r}
seqinr::choosebank("swissprot")
mySeq <- seqinr::query("mySeq", "N=MBP1_YEAST")
mbp1 <- seqinr::getSequence(mySeq)
seqinr::closebank()
x <- seqinr::AAstat(mbp1[[1]])
barplot(sort(x$Compo), cex.names = 0.6)
```
The same but now done via loading seqinr first, thus no need to reference the source package:
```{r}
library(seqinr)
choosebank("swissprot")
mySeq <- query("mySeq", "N=MBP1_YEAST")
mbp1 <- getSequence(mySeq)
closebank()
x <- AAstat(mbp1[[1]])
barplot(sort(x$Compo), cex.names = 0.6)
```
This is to avoid overhead at the beggining of a script in case the needed package is already installed:
```{r}
if (! requireNamespace("seqinr", quietly=TRUE)) {
  install.packages("seqinr")
}
```

Basic information on seqinr:
```{r}
library(help = seqinr)
```

Here it would show vignettes, if available
```{r}
browseVignettes("seqinr")
```

Showing available datasets for seqinr package
```{r}
data(package = "seqinr") 
```


```{r}
plot(cars)
```

```{r}
plot(cars)
```
